---
title: "hw4"
author: "yang an"
date: "`r Sys.Date()`"
output: html_document
---
## question
###answer:
Homework 4
```{r, echo=FALSE}

# Load the data 
setwd('/Users/yangan/Desktop/M236/M236-Lab/hw4')

options(scipen = 999)

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lme4)

# Load the Dental data
dental_data <- read.table("Dental.txt", header = TRUE, sep = "\t")
```
####1.
# (a)
```{r, echo=FALSE}
Xi_a <- matrix(1, nrow = 8, ncol = 1)  # 4 observations each for boy and girl
Xi_a
```
# (b)
```{r, echo=FALSE}
Xi_b <- matrix(c(1, 0, 1, 0, 1, 0, 1, 0,  # Boy's data
                 1, 1, 1, 1, 1, 1, 1, 1), # Girl's data
               nrow = 8, byrow = TRUE)
Xi_b
```
# (c)
```{r, echo=FALSE}
times <- c(8, 10, 12, 14)  # Same times for both boy and girl
Xi_c <- rbind(cbind(1, times), cbind(1, times))  # Combine both individuals
Xi_c
```
# (d)
```{r, echo=FALSE}
Xi_d <- rbind(
  cbind(1, times, 0),  # Boy's data (sex = 0)
  cbind(1, times, 1)   # Girl's data (sex = 1)
)
Xi_d
```
# (e)
```{r, echo=FALSE}
times <- c(8, 10, 12)  # Adjusted times
Xi_e <- rbind(
  cbind(1, times, 0),  # Boy's data
  cbind(1, times, 1)   # Girl's data
)
Xi_e
```
# (f)
```{r, echo=FALSE}
Xi_f <- rbind(
  cbind(1, times, 0, times^2, 0),  # Boy's data
  cbind(1, times, 1, times^2, 1)   # Girl's data
)
Xi_f
```
# (g)

We can use the same quadratic time trend as the baseline model, compare it with alternative models, and evaluate the significance of any improvement in fit and alignment with the data.

#####(h) 
```{r, echo=FALSE}
# Fit the models
# Load the package
library(lme4)

# (1a) No time trend, no gender effect
model_1a <- lmer(distance ~ 1 + (1 | id), data = dental_data)

# (1b) Different level by gender, no time trend
model_1b <- lmer(distance ~ gender + (1 | id), data = dental_data)

# (1c) Time trend, same for both groups
model_1c <- lmer(distance ~ age + (1 | id), data = dental_data)

# (1d) Time trend different for each group
model_1d <- lmer(distance ~ age * gender + (1 | id), data = dental_data)

# Summary of models
summary(model_1a)
summary(model_1b)
summary(model_1c)
summary(model_1d)
```
#####(i) 
1.(Intercept) (Estimate = 16.3406):
This is the estimated starting value (baseline distance) for males when age is 0. The average distance for males is 16.34 mm when age equal to 0.

2.age (Estimate = 0.7844):

This represents the average increase in distance per year of age for males (the reference gender). The distance increases by 0.784 mm per year for males.

3.genderGirl (Estimate = 1.0321):

This is the estimated difference in the baseline distance between females and males when age is 0.  Females is 1.032 mm greater than males when age is 0.

4.age:genderGirl
(Estimate = -0.3048):
This is the interaction term between age and gender (specifically for females), representing the difference in the slope of age for females compared to males. The slope of age for females is 0.3048 mm less than that for males.  Females have a slightly lower slope of 0.784âˆ’0.3048â‰ˆ0.479 mm per year while males have a slope of approximately 0.784 mm per year.

Summary of Fixed Effects
Baseline (age 0): Males start at an average distance of 16.34 mm, while females start at a slightly higher distance of 17.37 mm (16.34 + 1.032).
Time trend (slope): Males increase by 0.784 mm per year, while females increase by approximately 0.479 mm per year, indicating a smaller rate of increase in distance for females.
####2. 
#####(a) 
```{r, echo=FALSE}
Xi_unstructured_1 <- matrix(c(1, 0, 0, 0, 0,
                              1, 1, 0, 0, 0,
                              1, 1, 1, 0, 0,
                              1, 1, 1, 1, 0,
                              1, 1, 1, 1, 1), byrow = TRUE, ncol = 5)
Xi_unstructured_1
```
#####(b)

```{r, echo=FALSE}
Xi_unstructured_2 <- diag(5)
Xi_unstructured_2
```

#####(c) 
Five columns in both parts.
#####(d) 
In the first parametrization, each ð›¼ð‘˜represents the cumulative mean up to the k-th time. In the second, each ð›¼ð‘˜represents the mean for the k-th time point directly.

####3. 
#####(a) Fit each model separately. Report the fixed effects (regression output) results for all
```{r, echo=FALSE}
# Load necessary packages
library(lme4)
library(dplyr)

# Fit the models
model_1a <- lmer(distance ~ 1 + (1 | id), data = dental_data)
model_1b <- lmer(distance ~ gender + (1 | id), data = dental_data)
model_1c <- lmer(distance ~ gender + age + (1 | id), data = dental_data)
model_1d <- lmer(distance ~ age * gender + (1 | id), data = dental_data)
model_1f <- lmer(distance ~ age * gender + I(age^2) * gender + (1 | id), data = dental_data)

# Function to extract summary information for fixed effects
extract_fixed_effects <- function(model, model_name) {
  summary_data <- summary(model)$coefficients
  data.frame(
    Model = model_name,
    Term = rownames(summary_data),
    Estimate = summary_data[, "Estimate"],
    Std_Error = summary_data[, "Std. Error"],
    t_value = summary_data[, "t value"],
    stringsAsFactors = FALSE
  )
}

# Extract fixed effects for each model
model_1a_summary <- extract_fixed_effects(model_1a, "Model 1a")
model_1b_summary <- extract_fixed_effects(model_1b, "Model 1b")
model_1c_summary <- extract_fixed_effects(model_1c, "Model 1c")
model_1d_summary <- extract_fixed_effects(model_1d, "Model 1d")
model_1f_summary <- extract_fixed_effects(model_1f, "Model 1f")

# Combine summaries into one table
model_table <- bind_rows(model_1a_summary, model_1b_summary, model_1c_summary, model_1d_summary, model_1f_summary)

# Display the table
print(model_table)
```



#####(b) 
```{r, echo=FALSE}
# Extract AIC values for each model
model_aic <- c(AIC(model_1a), AIC(model_1b), AIC(model_1c), AIC(model_1d), AIC(model_1f))

# Find the model with the lowest AIC
best_model <- which.min(model_aic)

# Display the AIC values and best model
print(model_aic)
print(best_model)

# Extract the best model
best_model_summary <- extract_fixed_effects(get(paste0("model_1", letters[best_model])), paste0("Model 1", letters[best_model]))

# Display the summary of the best model
print(best_model_summary)

```
#####(c) 
Model 1d strikes the ideal balance by effectively capturing the significant influence of gender, age, and their interactions on distance, while preserving simplicity and avoiding unnecessary complexity.

#####(d) 
H0: There is no effect of gender or age:gender on the response variable distance.
```{r, echo=FALSE}
# Perform an F-test for the joint significance
anova(model_1c, model_1d)
```
The P-value of the F-test is less than 0.0006266<0.05, indicating that we reject the null hypothesis and conclude that both gender and the age:gender interaction are important predictors in the model.
  
####4.
Draw an inference plot. Estimate the average response and its standard error for the boys and the girls at each time point and plot over time.
Explain the difference between this inference plot and the empirical summary plot you drew for homework 3.
```{r, echo=FALSE}
# Load necessary packages
library(lme4)
library(ggplot2)
library(dplyr)

# Fit the model with age, gender, and interaction term
model <- lmer(distance ~ age * gender + (1 | id), data = dental_data)

# Use the original dataset for predictions, assuming all age-gender combinations are present
dental_data$fit <- predict(model, re.form = NA, se.fit = TRUE)$fit
dental_data$se.fit <- predict(model, re.form = NA, se.fit = TRUE)$se.fit

# Plot using the original dataset
ggplot(dental_data, aes(x = age, y = fit, color = gender, group = gender)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = fit - se.fit, ymax = fit + se.fit), width = 0.2) +
  labs(title = "Inference Plot for Boys and Girls", x = "Age (Years)", y = "Predicted Distance (mm)") +
  theme_minimal()


```
Both plots show the relationship between age and distance, but the inference plot includes the standard error of the predicted values, the confidence intervals is smaller than the empirical summary plot. The epirical summary plot shows the raw data points and the average distance at each time point, without any model assumptions, while the inference plot shows the predicted average distance with confidence intervals based on the model.

####5. 
#####(a) 
The most complicated fixed effect model is one that includes all possible interactionsï¼š
gender and age, gender and quadratic age. In addtion, the model also would include the gender 
effects, time trend, and the quadratic time trend.
#####(b) Fit the most complicated model:
```{r, echo=FALSE}
# Fit the most complex model
model_complex <- lmer(distance ~ age + gender + I(age^2) + age:gender + I(age^2):gender + (1 | id), data = dental_data)
summary(model_complex)
```
#####(c) Compare the two models using an F-test:
H0: The simpler model(best from earlier) is sufficient, and adding the quadratic terms does not significantly improve the fit.
H1: The complex model provides a significantly better fit to the model.
```{r, echo=FALSE}
# Perform an F-test to compare the two models
anova(model_1d, model_complex)
```
The P-value is 0.4906>0.05, we cannot reject the h0 and conclude that there is no significant impact to improve the model by adding the quadratic terms. Therefore, the simpler model is sufficient and should be preferred over the more complex model.

